---
title: "Fishnet Statistics — GLMM Fire × HAB Presence"
output: html_notebook
---

## Setup — Load Libraries and Data

```{r setup, message=FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(glmmTMB)
library(broom.mixed)
library(DHARMa)
library(ggplot2)
library(spaMM)
library(lme4)
library(ggeffects)
library(scales)
```

```{r load-data}
# Load main fishnet dataset
fishdata <- read_csv("FullFishnetdata.csv")

# Rename flag columns to descriptive names
names(fishdata)[names(fishdata) == "Fire_Flag"] <- "Fire_Presence"
names(fishdata)[names(fishdata) == "HAB_Flag"]  <- "HAB_Presence"

# Ensure grouping variables are factors
fishdata$Grid_ID  <- as.factor(fishdata$Grid_ID)
fishdata$Year     <- as.factor(fishdata$Year)
fishdata$Grid_Year <- as.factor(fishdata$Grid_Year)

head(fishdata)
```

```{r load-coords}
# Load fishnet dataset with spatial coordinates
fishandcoords <- read_csv("Fishnetandcoords.csv")

names(fishandcoords)[names(fishandcoords) == "Fire_Flag"] <- "Fire_Presence"
names(fishandcoords)[names(fishandcoords) == "HAB_Flag"]  <- "HAB_Presence"

fishandcoords$Year <- as.factor(fishandcoords$Year)
```

---

## Year × Fire/HAB Crosstabs

```{r crosstabs}
table(fishdata$Year, fishdata$Fire_Presence)
table(fishdata$Year, fishdata$HAB_Presence)
```

---

## GLMM Models

### Model 1 — Random intercepts for Grid_ID and Year

```{r model1}
model1 <- glmmTMB(HAB_Presence ~ Fire_Presence + (1 | Grid_ID) + (1 | Year),
                  data   = fishdata,
                  family = binomial)

summary(model1)
```

### Model 2 — Random intercept for Grid_Year interaction

```{r model2}
model2 <- glmmTMB(HAB_Presence ~ Fire_Presence + (1 | Grid_Year),
                  data   = fishdata,
                  family = binomial)

summary(model2)
```

### Model 3 — Model 2 with BFGS optimizer

```{r model3}
model3 <- glmmTMB(HAB_Presence ~ Fire_Presence + (1 | Grid_Year),
                  data    = fishdata,
                  family  = binomial,
                  control = glmmTMBControl(optimizer = optim,
                                           optArgs   = list(method = "BFGS")))

summary(model3)
```

### Null Model — Likelihood Ratio Test vs Model 1

```{r null-model}
model_null <- glmmTMB(HAB_Presence ~ (1 | Grid_ID) + (1 | Year),
                      data   = fishdata,
                      family = binomial)

anova(model_null, model1)
```

---

## Odds Ratios and Confidence Intervals

```{r odds-ratios}
confint_fixed  <- confint(model1, parm = "beta_", method = "Wald")
exp_confint    <- exp(confint_fixed)
or_estimates   <- exp(fixef(model1)$cond)

results <- data.frame(
  Term     = names(or_estimates),
  OR       = or_estimates,
  CI_lower = exp_confint[, 1],
  CI_upper = exp_confint[, 2]
)

print(results)
```

---

## Lagged Year Model

```{r lag-model}
# Create numeric year and lag by 1 year
fishdata$Year_numeric <- as.numeric(as.character(fishdata$Year))
fishdata$Year_Lag1    <- fishdata$Year_numeric - 1

model_lag_year <- glmmTMB(HAB_Presence ~ Fire_Presence + (1 | Grid_ID) + (1 | Year_Lag1),
                           data   = fishdata,
                           family = binomial)

summary(model_lag_year)
```

### AIC / BIC Comparison — Model 1 vs Lag Model

```{r model-comparison}
AIC(model1, model_lag_year)
BIC(model1, model_lag_year)
```

---

## DHARMa Residual Diagnostics

```{r dharma}
sim1    <- simulateResiduals(model1)
plot(sim1)

sim_lag <- simulateResiduals(model_lag_year)
plot(sim_lag)
```

---

## Fixed Effects Logistic Regression

```{r fixed-model}
model_fixed <- glm(HAB_Presence ~ Fire_Presence,
                   family = binomial,
                   data   = fishandcoords)

summary(model_fixed)
```

### Predicted Probability Plot — Fixed Model

```{r predicted-prob-plot}
newdata <- data.frame(Fire_Presence = c(0, 1))
newdata$predicted_prob <- predict(model_fixed, newdata, type = "response")
newdata$percent_label  <- paste0(round(newdata$predicted_prob * 100), "%")

ggplot(newdata, aes(x = factor(Fire_Presence), y = predicted_prob)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = percent_label), vjust = -0.5, size = 5) +
  scale_x_discrete(labels = c("No Fire", "Fire")) +
  scale_y_continuous(labels = percent_format(accuracy = 1),
                     limits = c(0, 0.20)) +
  ylab("Predicted Probability of HAB Presence") +
  xlab(NULL) +
  ggtitle("Fixed Effect Model Results") +
  theme_minimal()
```

---

## Spatial Model — Matérn Covariance

```{r spatial-model}
modela <- fitme(HAB_Presence ~ Fire_Presence + (1 | Year) +
                  Matern(1 | xcoord + ycoord),
                family = binomial(),
                data   = fishandcoords)

summary(modela)
```

---

## Mixed Effects Predicted Probabilities

```{r mixed-predictions}
model_mixed <- glmer(HAB_Presence ~ Fire_Presence + (1 | Grid_ID) + (1 | Year),
                     family = binomial(link = "logit"),
                     data   = fishdata)

predictions <- ggpredict(model_mixed, terms = "Fire_Presence")
predictions$percent_label <- paste0(round(predictions$predicted * 100), "%")

ggplot(predictions, aes(x = factor(x), y = predicted)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = percent_label), vjust = -0.5, size = 5) +
  scale_x_discrete(labels = c("No Fire", "Fire")) +
  scale_y_continuous(labels = percent_format(accuracy = 1),
                     limits = c(0, max(predictions$predicted) * 1.2)) +
  ylab("Predicted Probability of HAB Presence") +
  xlab("Wildfire Presence") +
  ggtitle("Predicted Probability of HAB Presence by Wildfire Status\n(Mixed-Effects Logistic Regression)") +
  theme_minimal()
```
