---
title: "PCA and Archetype Analysis — HAB Events by Lake Type"
output: html_notebook
---

## Setup — Load Libraries and Data

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)
library(grid)
library(scales)
library(FSA)
library(rstatix)
library(multcompView)
library(writexl)

load("8.17.25MLWork.RData")
```

---

## Combine HAB Data Across All Years

```{r combine-years}
totalHAB_data <- bind_rows(
  selected_data2019, selected_data2020,
  selected_data2021, selected_data2022, selected_data2023
)
```

---

## Join Lake Archetype Data

```{r join-archetypes}
# Export unique lake IDs (for reference)
dfg_ids <- totalHAB_data %>% distinct(DFGWATERID)
write_csv(dfg_ids, "DFGWATERID_list.csv")

# Load archetype classifications
lake_arch <- read_csv("CA_Lake_Arch_DFG.csv")

# Join arch column to HAB data
totalHAB_data <- totalHAB_data %>%
  left_join(lake_arch %>% select(DFGWATERID, arch),
            by = "DFGWATERID")

# Recode NAs and standardise HAB_Presence to numeric
totalHAB_data <- totalHAB_data %>%
  mutate(
    arch             = replace_na(arch, "Unknown"),
    HAB_Presence_num = if_else(HAB_Presence == "Yes", 1L, 0L)
  )
```

---

## PCA — Lake Morphological Variables

```{r pca-prep}
# Prepare numeric data for PCA (exclude arch_num from clustering variable)
df_pca_full <- totalHAB_data %>%
  mutate(arch_num = as.numeric(gsub("[^0-9]", "", arch))) %>%
  select(Wshd_area, Slope_100, Res_time, Lake_area, Vol_total,
         Shore_len, Depth_avg, arch, arch_num) %>%
  na.omit()

df_pca_numeric <- df_pca_full %>%
  select(Wshd_area, Slope_100, Res_time, Lake_area,
         Vol_total, Shore_len, Depth_avg)

# Run PCA
pca_res <- prcomp(df_pca_numeric, center = TRUE, scale. = TRUE)
summary(pca_res)
pca_res$rotation

# Variance explained
var_exp <- pca_res$sdev^2 / sum(pca_res$sdev^2)
pc1_var <- round(var_exp[1] * 100, 1)
pc2_var <- round(var_exp[2] * 100, 1)
```

```{r scree-plot}
var_expl <- var_exp
ggplot(data.frame(PC = seq_along(var_expl), Var = var_expl),
       aes(x = PC, y = Var)) +
  geom_line() + geom_point() +
  xlab("Principal Component") +
  ylab("Proportion of Variance Explained") +
  ggtitle("Scree Plot") +
  theme_minimal()
```

```{r pca-scores-loadings}
# Scores
scores          <- as.data.frame(pca_res$x)
scores$arch     <- df_pca_full$arch
scores$arch_num <- as.factor(df_pca_full$arch_num)

# Loadings with arrow scaling
loadings           <- as.data.frame(pca_res$rotation)
loadings$var       <- rownames(loadings)
arrow_scale        <- 5
loadings$PC1_arrow <- loadings$PC1 * arrow_scale
loadings$PC2_arrow <- loadings$PC2 * arrow_scale
```

### Biplot — Coloured by Archetype (filled points)

```{r biplot}
# Consistent archetype colours and line types used throughout
custom_colors <- c(
  "A1" = "#1b9e77", "A2" = "#d95f02", "A3" = "#7570b3",
  "A4" = "#e7298a", "A5" = "#66a61e", "A6" = "#e6ab02",
  "A7" = "#a6761d"
)
custom_linetypes <- c(
  "A1" = "dashed",  "A2" = "dotted",   "A3" = "dotdash",
  "A4" = "longdash","A5" = "twodash",  "A6" = "dashed",
  "A7" = "dotted"
)

ggplot(scores, aes(x = PC1, y = PC2, fill = arch_num)) +
  geom_point(shape = 21, color = "black",
             size = 3, stroke = 0.4, alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_segment(data = loadings,
               aes(x = 0, y = 0,
                   xend = PC1_arrow, yend = PC2_arrow),
               arrow = arrow(length = unit(0.2, "cm")),
               color = "gray30", inherit.aes = FALSE) +
  geom_text(data = loadings,
            aes(x = PC1_arrow, y = PC2_arrow, label = var),
            color = "black", size = 3,
            hjust = 0.5, vjust = -0.5, inherit.aes = FALSE) +
  scale_fill_brewer(palette = "Dark2") +
  coord_cartesian(xlim = c(-3, 3), ylim = c(-2.5, 2.5)) +
  labs(title = "PCA of Lake Variables (PC1 vs PC2)",
       x = paste0("PC1 (", pc1_var, "% variance)"),
       y = paste0("PC2 (", pc2_var, "% variance)"),
       fill = "Arch Num") +
  theme_minimal() +
  theme(legend.position = "right")
```

### Variable Contributions to PC1 and PC2

```{r variable-contributions}
var_contrib <- as.data.frame(pca_res$rotation^2) %>%
  mutate(var = rownames(.),
         across(where(is.numeric), ~ .x * 100))

# Separate ordered bar charts for PC1 and PC2
pc1_contrib <- var_contrib %>%
  select(var, PC1) %>%
  arrange(desc(PC1)) %>%
  mutate(var = factor(var, levels = rev(var)))

pc2_contrib <- var_contrib %>%
  select(var, PC2) %>%
  arrange(desc(PC2)) %>%
  mutate(var = factor(var, levels = rev(var)))

max_contrib <- max(c(pc1_contrib$PC1, pc2_contrib$PC2))

p1 <- ggplot(pc1_contrib, aes(x = var, y = PC1)) +
  geom_col(fill = "#1b9e77") +
  coord_flip(ylim = c(0, max_contrib)) +
  labs(title = "Variable Contributions to PC1",
       x = "Variable", y = "Contribution (%)") +
  theme_minimal()

p2 <- ggplot(pc2_contrib, aes(x = var, y = PC2)) +
  geom_col(fill = "#d95f02") +
  coord_flip(ylim = c(0, max_contrib)) +
  labs(title = "Variable Contributions to PC2",
       x = "Variable", y = "Contribution (%)") +
  theme_minimal()

p1
p2
```

---

## Archetype Summary Statistics

```{r arch-summary}
arch_summary <- totalHAB_data %>%
  filter(arch != "Unknown") %>%
  group_by(arch, DFGWATERID) %>%
  summarise(HAB_events = sum(HAB_Presence_num, na.rm = TRUE),
            .groups = "drop") %>%
  group_by(arch) %>%
  summarise(
    n_lakes             = n_distinct(DFGWATERID),
    total_HABs          = sum(HAB_events),
    mean_HABs_per_lake  = mean(HAB_events),
    prop_lakes_impacted = mean(HAB_events > 0),
    .groups = "drop"
  )

arch_summary
```

---

## Statistical Tests — HAB Events by Archetype

```{r kruskal-dunn}
lake_hab_events <- totalHAB_data %>%
  filter(arch != "Unknown") %>%
  group_by(DFGWATERID, arch) %>%
  summarise(HAB_events = sum(HAB_Presence_num, na.rm = TRUE),
            .groups = "drop") %>%
  mutate(log_HAB = log10(HAB_events + 1))

# Kruskal-Wallis
kruskal_res <- kruskal_test(HAB_events ~ arch, data = lake_hab_events)
print(kruskal_res)

# Dunn post-hoc (Bonferroni)
dunn_res <- lake_hab_events %>%
  dunn_test(HAB_events ~ arch, p.adjust.method = "bonferroni")
print(dunn_res)
```

### Chi-squared Test — HAB Presence by Archetype

```{r chisq}
lake_hab_binary <- totalHAB_data %>%
  filter(arch != "Unknown") %>%
  group_by(DFGWATERID, arch) %>%
  summarise(HAB_any = as.integer(any(HAB_Presence == "Yes")),
            .groups = "drop")

tab <- table(lake_hab_binary$arch, lake_hab_binary$HAB_any)
print(tab)
chisq.test(tab)
```

### Logistic Regression — HAB Presence ~ Archetype

```{r logistic}
fit <- glm(HAB_any ~ arch, data = lake_hab_binary, family = binomial)
summary(fit)
```

---

## Plots — HAB Events by Archetype

```{r boxplot-hab-events}
lake_counts <- lake_hab_events %>%
  group_by(arch) %>%
  summarise(n_lakes = n(), .groups = "drop")

ggplot(lake_hab_events, aes(x = arch, y = HAB_events + 1)) +
  geom_boxplot(fill = "skyblue", color = "black") +
  scale_y_log10() +
  geom_text(data = lake_counts,
            aes(x = arch, y = 1, label = paste0("n=", n_lakes)),
            inherit.aes = FALSE, vjust = -0.5, size = 3) +
  labs(x = "Lake Archetype",
       y = "Number of HAB Events (log scale)",
       title = "HAB Events by Lake Archetype") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r barplot-proportion}
hab_proportion <- lake_hab_events %>%
  group_by(arch) %>%
  summarise(
    n_lakes          = n(),
    n_lakes_with_HAB = sum(HAB_events > 0),
    prop_lakes_with_HAB = n_lakes_with_HAB / n_lakes,
    .groups = "drop"
  )

ggplot(hab_proportion, aes(x = arch, y = prop_lakes_with_HAB)) +
  geom_col(fill = "#1f77b4") +
  geom_text(aes(label = percent(prop_lakes_with_HAB, accuracy = 1)),
            vjust = -0.5) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(x = NULL, y = "Proportion of Lakes with HABs") +
  theme_minimal()
```

---

## Temporal Trends by Archetype

```{r temporal-prep}
arch_lakes <- totalHAB_data %>%
  filter(arch != "Unknown") %>%
  group_by(arch) %>%
  summarise(n_lakes = n_distinct(DFGWATERID), .groups = "drop")

hab_events_by_year <- totalHAB_data %>%
  filter(arch != "Unknown") %>%
  group_by(arch, Year) %>%
  summarise(
    total_HAB_events = sum(HAB_Presence_num, na.rm = TRUE),
    n_lakes_with_HAB = n_distinct(DFGWATERID[HAB_Presence_num > 0]),
    .groups = "drop"
  ) %>%
  left_join(arch_lakes, by = "arch") %>%
  mutate(prop_lakes_with_HAB = n_lakes_with_HAB / n_lakes)
```

```{r plot-proportion-trend}
ggplot(hab_events_by_year,
       aes(x = Year, y = prop_lakes_with_HAB,
           color = arch, group = arch, linetype = arch)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_color_manual(values = custom_colors) +
  scale_linetype_manual(values = custom_linetypes) +
  labs(y = "Proportion of Lakes with HABs",
       x = NULL, color = "Archetype", linetype = "Archetype") +
  theme_minimal() +
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r plot-total-events-trend}
ggplot(hab_events_by_year,
       aes(x = Year, y = total_HAB_events,
           color = arch, group = arch, linetype = arch)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_y_log10() +
  scale_color_manual(values = custom_colors) +
  scale_linetype_manual(values = custom_linetypes) +
  labs(x = "Year", y = "Total HAB Events (log scale)",
       title = "Temporal Trend of Total HAB Events by Lake Archetype",
       color = "Archetype", linetype = "Archetype") +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1))
```

---

## Save

```{r save}
write_xlsx(hab_events_by_year, "HAB_summary_by_arch_year.xlsx")
save.image(file = "ArchetypeHABs.RData")
message("✔ Saved")
```
